Viral Sequence Sorting Tools Evaluation

### 2022-11-01 EnvironmentalData/DrinkingWaterMetagenomes/building_env_data_viral_features_set_dw.Rmd
Visualizing viral totals by assembly.
Drinking water rule stats.

get rid of NAs for downstream processing
```{r}

# checkV
viruses$percent_viral[is.na(viruses$percent_viral)] <- 0
viruses$percent_unknown[is.na(viruses$percent_unknown)] <- 0
viruses$checkv_completeness[is.na(viruses$checkv_completeness)] <- 0

# VIBRANT
viruses$vibrant_quality[is.na(viruses$vibrant_quality)] <- 0

# DeepVirFinder
viruses$score[is.na(viruses$score)] <- 0
viruses$bh_pvalue[is.na(viruses$bh_pvalue)] <- 0

# VirSorter2
viruses$viral[is.na(viruses$viral)] <- 0
viruses$hallmark[is.na(viruses$hallmark)] <- 0

#Virsorter
viruses$category[is.na(viruses$category)] <- 0

# Kaiju
viruses$Kaiju_Viral[is.na(viruses$Kaiju_Viral)] <- "unknown"
```

@@

Data cleaning is done! Export viruses to tsv and load into identifying_viral_metrics.Rmd for metrics on tool outputs. The code for metrics is also pasted below, but it is not recommended running it in the same R session due to memory issues. The construction of the final dataframe essentially doubles the memory being used, and even with garbage collector can make Rstudio unresponsive.

>>>>

get rid of NAs for downstream processing
```{r}
# checkV
viruses$percent_viral[is.na(viruses$percent_viral)] <- 0
viruses$percent_unknown[is.na(viruses$percent_unknown)] <- 0
viruses$checkv_completeness[is.na(viruses$checkv_completeness)] <- 0
# VIBRANT
viruses$vibrant_quality[is.na(viruses$vibrant_quality)] <- 0
# DeepVirFinder
viruses$score[is.na(viruses$score)] <- 0
viruses$bh_pvalue[is.na(viruses$bh_pvalue)] <- 0
# VirSorter2
viruses$viral[is.na(viruses$viral)] <- 0
viruses$hallmark[is.na(viruses$hallmark)] <- 0
viruses$RNA <- 0
viruses$RNA[is.na(viruses$RNA)] <- 0
viruses$lavidaviridae <- 0
viruses$lavidaviridae[is.na(viruses$lavidaviridae)] <- 0
viruses$NCLDV <- 0
viruses$NCLDV[is.na(viruses$NCLDV)] <- 0
viruses$ssDNA[is.na(viruses$ssDNA)] <- 0
viruses$dsDNAphage[is.na(viruses$dsDNAphage)] <- 0
viruses$viral[is.na(viruses$max_score)] <- 0
#Virsorter
viruses$category[is.na(viruses$category)] <- 0
# Kaiju
viruses$Kaiju_Viral[is.na(viruses$Kaiju_Viral)] <- "unknown"
viruses$Kingdom[is.na(viruses$Kingdom)] <- "unknown"
#adding a combined provirus category
viruses$provirus <- FALSE
viruses$provirus[viruses$checkv_provirus=="Yes" | viruses$vibrant_prophage=="Yes" | viruses$category==4] <- TRUE
```

@@

Data cleaning is done! Export viruses to tsv and load into identifying_viral_contigs_environmental.Rmd for metrics on tool outputs. The code for metrics is also pasted below, but it is not recommended running it in the same R session due to memory issues. The construction of the final dataframe essentially doubles the memory being used, and even with garbage collector can make Rstudio unresponsive.

### 2022-11-01 EnvironmentalData/DrinkingWaterMetagenomes/identifying_viral_contigs_environmental_dw.Rmd
Drinking water rule stats.

    if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="complete circular"] <- keep_score[input_seqs$vibrant_quality=="complete circular"] + 1
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$checkv_provirus=="Yes"] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$checkv_provirus=="Yes"] + 0.5
  }

  if (include_virsorter2) {
    keep_score[input_seqs$viral>=50] <- keep_score[input_seqs$viral>=50] + 0.5
    keep_score[input_seqs$viral>=95] <- keep_score[input_seqs$viral>=95] + 0.5
    keep_score[input_seqs$RNA>=0.9] <- keep_score[input_seqs$RNA>=0.9] + 1
    keep_score[input_seqs$lavidaviridae>=0.9] <- keep_score[input_seqs$lavidaviridae>=0.9] + 1
    keep_score[input_seqs$NCLDV>=0.9] <- keep_score[input_seqs$NCLDV>=0.9] + 1
  }

@@

  if (include_tuning_viral) {
    keep_score[input_seqs$Kaiju_Viral=="Viruses"] <- keep_score[input_seqs$Kaiju_Viral=="Viruses"] + 0.5
    keep_score[input_seqs$hallmark>2] <- keep_score[input_seqs$hallmark>2] + 1
    keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] <- keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] + 0.5
    keep_score[input_seqs$percent_viral>=50] <- keep_score[input_seqs$percent_viral>=50] + 0.5
  }

  if (include_tuning_not_viral) {
    keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] <- keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] - 0.5
    keep_score[input_seqs$checkv_host_genes>50 & input_seqs$checkv_provirus=="No"] <- keep_score[input_seqs$checkv_host_genes>50 & input_seqs$checkv_provirus=="No"] - 1
    keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] <- keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] - 1
    keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & input_seqs$checkv_provirus=="No"] <- keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & input_seqs$checkv_provirus=="No"] - 1 # consider accounting for RNA viruses
    keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] <- keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] - 1
    keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] <- keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] - 0.5 # helped with protist contamination
  }

@@

>>>>

  if (include_vibrant) {
    keep_score[input_seqs$vibrant_quality=="complete circular"] <- keep_score[input_seqs$vibrant_quality=="complete circular"] + 1
    keep_score[input_seqs$vibrant_quality=="high quality draft"] <- keep_score[input_seqs$vibrant_quality=="high quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="medium quality draft"] <- keep_score[input_seqs$vibrant_quality=="medium quality draft"] + 1
    keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] <- keep_score[input_seqs$vibrant_quality=="low quality draft" & input_seqs$provirus] + 0.5
  }

  if (include_virsorter2) {
    keep_score[input_seqs$viral>=50 | input_seqs$lavidaviridae>=0.95 | input_seqs$NCLDV>=0.95] <- keep_score[input_seqs$viral>=50 | input_seqs$lavidaviridae>=0.95 | input_seqs$NCLDV>=0.95] + 0.5
    keep_score[input_seqs$viral>=95] <- keep_score[input_seqs$viral>=95] + 0.5
    keep_score[input_seqs$RNA>=0.95] <- keep_score[input_seqs$RNA>=0.95] + 1
  }

@@

  if (include_tuning_viral) {
    keep_score[input_seqs$Kaiju_Viral=="Viruses"] <- keep_score[input_seqs$Kaiju_Viral=="Viruses"] + 0.5
    keep_score[input_seqs$hallmark>2] <- keep_score[input_seqs$hallmark>2] + 0.5
    keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] <- keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] + 0.5
    keep_score[input_seqs$percent_viral>=50] <- keep_score[input_seqs$percent_viral>=50] + 0.5
  }

  if (include_tuning_not_viral) {
    keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] <- keep_score[input_seqs$Kaiju_Viral=="cellular organisms"] - 0.5
    keep_score[input_seqs$checkv_host_genes>50 & !input_seqs$provirus] <- keep_score[input_seqs$checkv_host_genes>50 & !input_seqs$provirus] - 1
    keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] <- keep_score[input_seqs$checkv_viral_genes==0 & input_seqs$checkv_host_genes>=1] - 1
    keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & !input_seqs$provirus] <- keep_score[((input_seqs$checkv_viral_genes*3) <= input_seqs$checkv_host_genes) & !input_seqs$provirus] - 1 
    keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] <- keep_score[input_seqs$checkv_length>500000 & input_seqs$hallmark<=1] - 1
    keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] <- keep_score[input_seqs$checkv_completeness<=75 & input_seqs$checkv_length<=5000] - 0.5 
  }

@@

# numbers in common between two tool combos
```{r}
num_high_viruses <- matrix(data=NA, nrow=63, ncol=63)

for (i in 1:63) {
  for (j in 1:63) {
    num_high_viruses[i,j] <- sum(viral_scores[,i]>=1 & viral_scores[,j]>=1)
  }
}

rownames(num_high_viruses) <- combos_list$toolcombo
colnames(num_high_viruses) <- combos_list$toolcombo
```

```{r}
combos_list_two <- combos_list
combos_list_two$numtools <- str_count(combos_list_two$toolcombo, "1")
combos_list_two <- combos_list_two[combos_list_two$numtools==1,]

two_tool_pairs <- num_high_viruses[rownames(num_high_viruses) %in% combos_list_two$toolcombo,
                                   colnames(num_high_viruses) %in% combos_list_two$toolcombo]

colnames(two_tool_pairs) <- c("VS2", "VS", "VB", "tv", "DVF", "tnv")
rownames(two_tool_pairs) <- c("VS2", "VS", "VB", "tv", "DVF", "tnv")
```

```{r}
viral_scores_two <- viral_scores[,colnames(num_high_viruses) %in% combos_list_two$toolcombo]

n <- 6
viral_scores_two_comb <- matrix(data=NA, nrow=n, ncol=n)

for (i in 1:n) {
  for (j in 1:n) {
    if (i==j) {
      viral_scores_two_comb[i,j] <- sum(viral_scores_two[,i]>=1)
    }
    else if (i<j) {
      viral_scores_two_comb[i,j] <- sum((viral_scores_two[,i]+viral_scores_two[,j])>=1)
    }
    else if (i>j) {
      viral_scores_two_comb[i,j] <- paste(round(sum((viral_scores_two[,i]+viral_scores_two[,j])>=1)/sum(viral_scores_two[,i]>=1 | viral_scores_two[,j]>=1)*100, digits=0), "%", sep="")
    }
  }
}

colnames(viral_scores_two_comb) <- c("VS2", "VS", "VB", "tv", "DVF", "tnv")
rownames(viral_scores_two_comb) <- c("VS2", "VS", "VB", "tv", "DVF", "tnv")
```

```{r}
viral_scores_two_comb
```

### 2022-11-01 TestingPipeline/Scripts/rule_tuning_2022.09.06.Rmd
Drinking water rule stats.

Tuning - Viral Signature:
    Kaiju_viral = "Viruses": +1
    If %unknown >= 75 and length < 50000: +0.5
    If %viral >= 50: +0.5
    Hallmark > 2: +1

@@

  if (include_tuning_viral) {
    keep_score[input_seqs$Kaiju_Viral=="Viruses"] <- keep_score[input_seqs$Kaiju_Viral=="Viruses"] + 0.5
    keep_score[input_seqs$hallmark>2] <- keep_score[input_seqs$hallmark>2] + 1
    keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] <- keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] + 0.5
    keep_score[input_seqs$percent_viral>=50] <- keep_score[input_seqs$percent_viral>=50] + 0.5
  }

@@

stats for different types of sequences
```{r}
TP <- table(viruses$confusion_matrix_high_MCC)[4]
FN <- table(viruses$confusion_matrix_high_MCC)[1]

recall <- as.numeric(TP/(TP+FN))
recall
```

```{r}
head(viruses$contig[viruses$seqtype=="virus"])
tail(viruses$contig[viruses$seqtype=="virus"])
```

>>>>

Tuning - Viral Signature:
    Kaiju_viral = "Viruses": +0.5
    If %unknown >= 75 and length < 50000: +0.5
    If %viral >= 50: +0.5
    Hallmark > 2: +0.5

@@

  if (include_tuning_viral) {
    keep_score[input_seqs$Kaiju_Viral=="Viruses"] <- keep_score[input_seqs$Kaiju_Viral=="Viruses"] + 0.5
    keep_score[input_seqs$hallmark>2] <- keep_score[input_seqs$hallmark>2] + 0.5
    keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] <- keep_score[input_seqs$percent_unknown>=75 & input_seqs$checkv_length<50000] + 0.5
    keep_score[input_seqs$percent_viral>=50] <- keep_score[input_seqs$percent_viral>=50] + 0.5
  }

@@
